Continuous estimates of heat emission at Mt. Ruapehu using the Unscented Kalman Smoother

### 2022-02-03 manuscript/manuscript.tex
update style file locations
Version 2 of the manuscript

%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
\documentclass{bmc_template/bmcart}

@@

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   corref={aff1},                       % id of corresponding address, if any
   noteref={n1},                        % id's of article notes, if any
   email={y.behr@gns.cri.nz}   % email address
]{\inits{YB}\fnm{Yannik} \snm{Behr}}
\author[
   addressref={aff1},
   email={s.sherburn@gns.cri.nz}
]{\inits{SS}\fnm{Steve} \snm{Sherburn}}
\author[
   addressref={aff1},
   email={t.hurst@gns.cri.nz}
]{\inits{TH}\fnm{Tony} \snm{Hurst}}

@@

\begin{abstract} % abstract
\parttitle{First part title} %if any
Text for this section.

\parttitle{Second part title} %if any
Text for this section.
\end{abstract}

@@

\begin{keyword}
\kwd{sample}
\kwd{article}
\kwd{author}
\end{keyword}

@@

Monitoring changes in temperature, water mass, and ion concentration in these
lakes can thus serve as proxies for changes in gas- and steam-release by the
hydrothermal system and, ultimately, the melt zone beneath the volcano.

@@

Following \cite{Hurst1991} and \cite{Stevenson1992} the mass and energy balance
can be written in terms of three coupled ordinary differential equations. The
rate of lake water mass ($M$) change can be described as:
\begin{equation}\label{ode_M}
	\frac{dM}{dt} = \dot{M_s} - \dot{M_o} - \dot{M_e} + \dot{M_i}
\end{equation}

@@

where $T$ is the lake water temperature, $\dot{Q_r}$ is the rate of energy gain
due to solar irradiation (short-wavelength radiation), and $c_wT\dot{M_o}$ is
the rate of energy loss through outflow/seepage. $\dot{Q_e}$ is the rate of
surface energy losses due to evaporation (forced and free convection), sensible
heat, and long-wavelength radiation (net loss). Several empirical and
theoretical equations exist to describe the surface losses and we refer the
reader to previous publications for more detailed discussions
\cite{Stevenson1992, hurstCraterLakeEnergy2015}. There is no consensus on the
best quantitative model of evaporation from warm lakes and, in fact, the most
suited equation may depend on the location of and available observations from
the lake. Appendix \ref{A} describes the equations used in this study in detail
which include the recent development on forced (i.e. wind-induced) convection
\cite{sartoriCriticalReviewEquations2000}.

@@

In a sensitivity analysis for $\dot{T}$ (Figure \ref{sens}) we can see a linear
relationship with respect to $\dot{Q_i}$ and $\dot{M_o}$ and a mildly
non-linear relationship with respect to $\dot{Q_i}$ and current season
(note that this assumes southern hemisphere seasons). The relationship with respect
to $T$ and windspeed, however, is strongly non-linear. This is because $\dot{Q_e}$
and $\dot{M_e}$ are non-linear functions of windspeed and $\dot{Q_i}$. This further 
complicates the inference of $\dot{Q_i}$.

\begin{figure}
    \includegraphics[width=12 cm]{figures/sensitivity_analysis.png}  
    \caption{Sensitivity analysis of $\dot{T}$ with respect to $\dot{Q_i}$ 
             and $\dot{M_o}$ (top panel), $\dot{Q_i}$ and season 
             (center panel), and $T$ and windspeed (bottom panel). The model
              parameters that have been held fixed are described in each 
              panel's title. Warmer colors indicate a temperature increase
              ($\dot{T} > 0$) and cooler colors a temperature decrease
              ($\dot{T} < 0$)}
      \label{sens}
\end{figure}

@@

where $\bm{y}_{1:t}$ is the time series of $m$ observations and $x_k$ comprises
the $n$ model parameters described in Section \ref{phm} at time step $k$. Note
that $x_k$ also includes $T$, $M$, $\dot{M_o}$, and $X$. We distinguish here
between the state of the lake and observations of this state
($T$, $M$, $X$, $\dot{M_o}$) which are to some degree uncertain.

Solving Equation \ref{bsm} using standard strategies like Markov Chain Monte
Carlo is computationally very expensive as the number of variables grows
exponentially with the length of the timeseries. An alternative are Bayesian
Filtering and Smoothing algorithms of which the Kalman Filter and Smoother are
the closed form solution for linear models \cite{Kalman1960, Rauch1965,
sarkkaBayesianFilteringSmoothing2013}.

@@

where $f(\bm{x}_{k-1})$ is the physical model described in Section \ref{phm}
propagated forward in time using a fourth-order Runge-Kutta scheme. If we 
assume the errors, $\bm{r}_k$ and $\bm{q}_{k}$, to be normally distributed
($\bm{q}_{k} \sim N(\bm{0},
\bm{Q}_{k})$, $\bm{r}_k \sim N(\bm{0}, \bm{R}_k)$), we can compute an
approximate solution to \ref{bsm1} using the \textit{Unscented Kalman Filter}
\citep{Merwe2004, sarkkaBayesianFilteringSmoothing2013}.

@@

To evaluate the inference scheme we design a synthetic test, consisting of a
very large cylinder, the same volume as RCL. We keep the windspeed and surface
water inflow constant at 4.5 m/s and 10 kt/day, respectively. To model the
outflow we assume the outlet to be a pipe 0.2 $m^2$ in cross section and then
use Bernoulli's equation to model the water outflow based on the water level in
the cylinder above the pipe. We assume that the incoming heat follows a
combination of a third-order polynomial and a sinusoid function and that the
enthalpy of the incoming steam is 2.8 MJ/kg which corresponds to the enthalpy of
saturated steam at hydrostatic pressures similar to those at the bottom of the
lake \cite{Mayhew1978}. To the resulting synthetic observations we add
uncertainties similar to those estimated from real observations (see Section
\ref{realdata}).

@@

It is noteworthy that the mass influx and outflux rates are recovered
quite accurately despite having no data in the former and very little data
in the latter case.

\begin{figure}
	\includegraphics[width=12 cm]{figures/synthetic_inversion_uks.png}  
    \caption{Inversion result for the synthetic test described in Section 
        \ref{syn_test}. Turquoise lines and markers represent hypothetical
        observations; red dashed lines are the inversion results and black
        dashed lines are the true input to the synthetic test.}
  \label{syn_example}
\end{figure}

\subsection{Real data}\label{realdata}
Temperature at RCL is measured by an integrated-circuit temperature device
(Texas Instruments LM35) located roughly 1.9 m below the lake's current
overflow level. The water level sensor is a hydrostatic pressure sensor (First
Sensor KTE/KTU/KTW6000\ldots CS Series) and it is co-located with the
temperature sensor. Both sensors have a sampling interval of 15 minutes and
transmit data several times a day via satellite to GNS Science, which monitors
New Zealand's geohazards through its GeoNet project (www.geonet.org.nz
 last accessed 16/07/2021).
Water samples are taken 8-10 times a year from which subsequently, among many
other analytes, concentrations of $Mg^{++}$ and $Cl^-$ are determined at GNS
Science's Geothermal Analytical Laboratory. We will focus here on $Mg^{++}$
concentration as it shows less deviation from the assumption of constant
dilution than $Cl^-$.

@@

Figure \ref{inference_result} shows $\dot{Q_i}$ for RCL for lake measurements
between 4 March 2016 and 7 July 2021. During this time the heat input rate
reached a maximum of 753 MW and was on average 165 MW. The average standard
deviation was 34 MW and generally increased for lower values of $\dot{Q_i}$. To
put this in context, currently installed geothermal power plants in New Zealand
have a combined output of $\sim$~1000MW.

\begin{figure}
 	\includegraphics[width=13 cm]{figures/uks_real_data.png}  
 	\caption{Inference results for the heat input rate $\dot{Q_i}$ at RCL
             between 4 March 2016 and 7 July 2021. The solid line shows the
             marginal probability and the uncertainty is displayed as shaded
             region.}
\label{inference_result}
\end{figure}

\section{Discussion and conclusion}
Inferring the heat input rate into a crater lake is an important part in
estimating the total energy budget of a volcanic system \cite{Brown1989a}.
Similar to a volcano's gas budget, it can either indicate migration of magma or
changes in permeability within the vent system.

Figure \ref{psd} shows the power spectral density (PSD) for the marginal heat
input rate shown in Figure \ref{inference_result}. It was computed using the
multi taper method \cite{prietoComputersGeosciencesFortran2009} which provides
not only an estimate of the PSD but also its uncertainty. The PSD shows that the
dominant periods in the heat input rate lie between $\sim$100 and $\sim$250
days. Whether these periods are due to cyclic processes in the magmatic system,
the hydrothermal system, or a combination of both requires further modelling of
these processes. Many studies have demonstrated the value of non-linear Kalman
Filters for inference and data-assimilation with non-linear, high-dimension
numerical models \cite{White2018a}. While beyond the scope of this study, the
next logical step, therefore, is to include numerical models of magma degassing
and hydrothermal fluid flow to model the heat input rate, volume, and enthalpy
of the steam entering the lake.

\begin{figure}
    \includegraphics[width=12 cm]{figures/psd_analysis.png}  
 	\caption{Power spectral density of the heat input rate 
             (Figure \ref{inference_result}) shown as a solid line
             with uncertainty displayed as shaded region.}
    \label{psd}
\end{figure}

Irrespective of the process that is responsible for the periodicity, it appears
that it has not changed significantly over the past five years. This may
suggest that the volcano has been in a dynamic equilibrium during this time.
This is further supported by the cumulative heat input (Figure \ref{cumheat}).
It shows a slight increase in heat input in mid-2017 but a fairly constant rate
since then. Of course this does not in any way forecast the volcano's future
behaviour but it may serve as a useful baseline to detect changes in the
volcano's state more easily.

\begin{figure}
    \includegraphics[width=12 cm]{figures/cumulative_heat.png}
    \caption{Cumulative heat input into RCL (solid line) and its 
    uncertainty (shaded region).}
    \label{cumheat}
\end{figure}

Previous studies of RCL's mass and energy balance found that they required very
high values of enthalpy for the incoming steam to avoid violating physical
boundary conditions such as negative values for the inflow rate of meteoric
water \cite{hurstCraterLakeEnergy2015, Stevenson1992}. Their explanation was the
existence of a heat pipe beneath the lake where gas and fluid phases coexist. As
the rising gas enters the lake relatively cool lake water enters the vent
leading to an overall lower addition of mass to the lake and resulting in an
enthalpy value that is double than expected for saturated steam at hydrostatic
pressures likely to be found at the bottom of the lake. By including
uncertainties in our observations we no longer require these high enthalpy
values to satisfy the physical constraints. Numerical modelling of hydrothermal
processes may also shed further light on the heating processes and whether or
not a heat pipe is a plausible mechanism.

>>>>

%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
\documentclass{bmcart}

@@

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   corref={aff1},                       % id of corresponding address, if any
   email={y.behr@gns.cri.nz}   % email address
]{\inits{YB}\fnm{Yannik} \snm{Behr}}
\author[
   addressref={aff1},
   email={s.sherburn@gns.cri.nz}
]{\inits{SS}\fnm{Steve} \snm{Sherburn}}
\author[
   addressref={aff1},
   email={t.hurst@gns.cri.nz}
]{\inits{TH}\fnm{Tony} \snm{Hurst}}

@@

\begin{abstract} % abstract
    Volcanic lakes often capture a significant amount of volcanic heat emission
    and thus provide a unique opportunity to monitor changes inside the volcano.
    We present a Bayesian inversion method to automatically infer volcanic heat
    emission at the base of a volcanic lake from lake monitoring data using a
    non-linear Kalman Smoother. Our method accounts for the, in parts large,
    uncertainties in observations and the underlying physics-based model to
    generate probabilistic estimates of heat emission. We verify our results
    using a synthetic test case and then estimate the daily heat input rate into
    Mt. Ruapehu's Crater Lake, New Zealand, between 2016 and 2022. Time-frequency
    analysis of the heat input rate shows dominant periods of heating cycles
    ranged between 100 - 250 days. Some years are dominated by shorter cycles
    and others by longer cycles which potentially points to changes in the
    magmatic heat supply and the hydrothermal system.
\end{abstract}

@@

\begin{keyword}
\kwd{Crater lakes}
\kwd{Non-linear Kalman Smoother}
\kwd{Volcanic heat emission}
\end{keyword}

@@

Monitoring changes in temperature, water mass, and ion concentration in these
lakes can, therefore, serve as proxies for changes in gas- and steam-release by
the hydrothermal system and, ultimately, the melt zone beneath the volcano.

@@

The mass and energy balance model can be written in terms of three coupled ordinary
differential equations \cite{Hurst1991, Stevenson1992}. The rate of lake water
mass ($M$) change is described as:
\begin{equation}\label{ode_M}
	\frac{dM}{dt} = \dot{M_s} - \dot{M_o} - \dot{M_e} + \dot{M_i}
\end{equation}

@@

where $T$ is the lake water temperature, $\dot{Q_r}$ is the rate of energy gain
due to solar irradiation (short-wavelength radiation), and $c_wT\dot{M_o}$ is
the rate of energy loss through outflow/seepage. $\dot{Q_e}$ is the rate of
surface energy losses due to evaporation (forced and free convection), sensible
heat, and long-wavelength radiation (net loss). Several empirical and
theoretical equations exist to describe the surface losses and we refer the
reader to previous publications for more detailed discussions
\cite{Stevenson1992, hurstCraterLakeEnergy2015}. There is no consensus on the
best quantitative model of evaporation from warm lakes and, in fact, the most
suited equation may depend on the location of and available observations from
the lake. Appendix \ref{A} describes the equations used in this study in detail
which include the recent development on forced (i.e. wind-induced) convection
\cite{sartoriCriticalReviewEquations2000}. It is important to note here that
$Q_e$ is a function of $T$ and the windspeed above the lake.

@@

The main challenge in solving the mass and energy balance equations for
$\dot{Q_i}$ arises from the non-linear relationship between $Q_e$ and ${T}$
as shown in Figure \ref{sens}.

\begin{figure}
    \includegraphics[width=12 cm]{figures/sensitivity_analysis.png}  
    \caption{Surface energy loss ($Q_e$) with respect to lake water temperature
             ($T$). The model parameters that have been held fixed are shown in  
             the title.}
      \label{sens}
\end{figure}

@@

where $\bm{y}_{1:t}$ is the time series of $m$ observations and $x_k$ comprises
the $n$ model parameters described in Section \ref{phm} at time step $k$. Note
that $x_k$ also includes $T$, $M$, and $X$. We distinguish here between the
state of the lake and observations of this state ($T$, $M$, $X$) which are to
some degree uncertain. Equation \ref{bsm} can be seen as the probabilistic
formulation of the inverse problem inferring $\dot{Q_i}$ from available
observations.

Solving Equation \ref{bsm} using standard strategies like Markov Chain Monte
Carlo is computationally very expensive as the number of variables grows
exponentially with the length of the timeseries and the number of model
parameters. An alternative are Bayesian Filtering and Smoothing algorithms of
which the Kalman Filter and Smoother are the closed form solution for linear
models \cite{Kalman1960, Rauch1965, sarkkaBayesianFilteringSmoothing2013}.

@@

where $f(\bm{x}_{k-1})$ is the non-linear physical model described in Section
\ref{phm} propagated forward in time using a fourth-order Runge-Kutta scheme. If
we assume the errors, $\bm{r}_k$ and $\bm{q}_{k}$, to be normally distributed
($\bm{q}_{k} \sim N(\bm{0}, \bm{Q}_{k})$, $\bm{r}_k \sim N(\bm{0}, \bm{R}_k)$),
we can compute an approximate solution to \ref{bsm1} using the \textit{Unscented
Kalman Smoother} \citep{Merwe2004, sarkkaBayesianFilteringSmoothing2013}, a
non-linear Kalman Smoother.

@@

To evaluate the inference scheme we design a synthetic test, consisting of a
very large cylinder, the same volume as RCL. We keep the windspeed and surface
water inflow constant at 4.5 m/s and 10 kt/day, respectively. To model the
outflow we assume the outlet to be a pipe 0.2 $m^2$ in cross section and then
use Bernoulli's equation to model the water outflow based on the water level in
the cylinder above the pipe. We assume that the incoming heat follows a
combination of a third-order polynomial and a sinusoid function and that the
enthalpy of the incoming steam is 2.8 MJ/kg which corresponds to the enthalpy of
saturated steam at hydrostatic pressures similar to those at the bottom of the
lake \cite{Mayhew1978}. To the resulting synthetic observations we add
uncertainties similar to those estimated from real observations (see Section
\ref{realdata}). As with real data, average and standard deviation of all values
within a day are calculated. The standard deviation is taken as a proxy for the
uncertainty in observations ($\bm{R}_k$). To simulate the disparity between sampling
intervals of different data streams we randomly remove most of the observations
of $X$, $\dot{M_o}$, and $\dot{M_s}$. This dataset is then inverted using the
Unscented Kalman Smoother to recover the original heat input rate.

@@

It is noteworthy that the mass outflux rates are recovered quite accurately
despite providing very little synthetic observations.

\begin{figure}
	\includegraphics[width=12 cm]{figures/synthetic_inversion_uks.png}  
    \caption{Inversion result for the synthetic test described in Section 
        \ref{syn_test}. Turquoise lines and markers represent synthetic 
        observations; red dashed lines are the inversion results and black
        dashed lines are the true input to the synthetic test. Error bars 
        represent $\pm$ 3 standard deviations.}
  \label{syn_example}
\end{figure}

\subsection{Real data}\label{realdata}
Temperature at RCL is measured by an integrated-circuit temperature device
(Texas Instruments LM35) located roughly 1.9 m below the lake's current
overflow level. The water level sensor is a hydrostatic pressure sensor (First
Sensor KTE/KTU/KTW6000\ldots CS Series) and it is co-located with the
temperature sensor. Both sensors have a sampling interval of 15 minutes and
transmit data several times a day via satellite to GNS Science, which monitors
New Zealand's geohazards through its GeoNet project \cite{GeoNetHome}.
Water samples are taken 8-10 times a year from which subsequently, among many
other analytes, concentrations of $Mg^{++}$ and $Cl^-$ are determined at GNS
Science's Geothermal Analytical Laboratory. We will focus here on $Mg^{++}$
concentration as it shows less deviation from the assumption of constant
dilution than $Cl^-$.

@@

Figure \ref{inference_result}a shows $\dot{Q_i}$ for RCL for lake measurements
between 4 March 2016 and 1 February 2022. During this time the heat input rate
reached a maximum of 753 MW and was on average 165 MW. The average standard
deviation was 34 MW and generally increased for lower values of $\dot{Q_i}$. To
put this in context, currently installed geothermal power plants in New Zealand
have a combined output of $\sim$~1000MW.

To also investigate changes in periodicity we computed the continuous wavelet
transform of of $\dot{Q_i}$ using a complex Morlet wavelet (Figure
\ref{inference_result}b). We chose the wavelet transform over the more common
windowed Fourier transform for its ability to detect non-stationary signals
\cite{torrencePracticalGuideWavelet1998}. Figure \ref{inference_result}c shows
the global wavelet spectrum which is an average of the wavelet analysis in
Figure \ref{inference_result}b over the whole analysis period. The global
wavelet spectrum can be interpreted similarly to the Power Spectral Density.
Figure \ref{inference_result} shows that periodicity is dominated by periods
between 100 and 250 days. Which period is dominant at any one time changes
significantly and we will expand on this further in the following section.

\begin{figure}
 	\includegraphics[width=12 cm]{figures/time_frequency_analysis.png}  
    \caption{ a) Inference results for the heat input rate $\dot{Q_i}$ at RCL
             between 4 March 2016 and 1 February 2022. The solid line shows the
             marginal probability and the uncertainty is displayed as shaded
             region; b) Continuous wavelet transform of the time series shown in
             Panel a; c) Power spectral density of the heat input rate shown as a
             solid line with uncertainty displayed as shaded region.}
\label{inference_result}
\end{figure}

\section{Discussion and conclusion}
Inferring the heat input rate into a crater lake is an important part in
estimating the total energy budget of a volcanic system \cite{Brown1989a}.
Similar to a volcano's gas budget, changes in the heat input rate can either
indicate changes in the magmatic heat source or in the hydrothermal system
above.

We would expect a significant change in the heat source to cause a change in
slope of the cumulative heat input rate (Figure \ref{cumulative}). The fact that
we do not observe such changes may indicate a more or less constant supply of
heat during the studied period. Experiments on analog models of hydrothermal
systems suggest that changes in the heat source and the hydrothermal system
should also change the length of heating cycles
\cite{vandemeulebrouckAnalogueModelingInstabilities2005,fitzgerald1997TwentySecondWorkshop}.
Figure \ref{inference_result}b shows the heating cycles before mid-2017 and
after mid-2019 were dominated by longer periods. From mid-2017 to mid-2019
cycles were mostly of shorter periods which may indicate a change in the
volcanic system during this time.

\begin{figure}
 	\includegraphics[width=12 cm]{figures/cumulative_heat.png}  
    \caption{Cumulative heat input into RCL (solid line) and its uncertainty 
             (shaded region) calculated for the time-series in
             Figure \ref{inference_result} a.} 
\label{cumulative}
\end{figure}

To further explore hypotheses like this, the next logical step is to include
numerical models of magma degassing and hydrothermal heat and fluid flow. While
beyond the scope of this study, our inversion method is well suited for such
more complex models as its runtime scales linearly with the number of
parameters. Several studies have demonstrated the value of non-linear Kalman
Filters for inverse problems with non-linear, high-dimensional numerical models
\cite[e.g.][]{White2018a,huangIteratedKalmanMethodology2022}.

Previous studies of RCL's mass and energy balance found that they required very
high values of enthalpy for the incoming steam to avoid violating physical
boundary conditions such as negative values for the inflow rate of meteoric
water ($\dot{M_s}$) \cite{hurstCraterLakeEnergy2015, Stevenson1992}. Their
explanation was that as rising hot fluids enter the lake, relatively cool lake
water flows back into the vent. This would lead to an overall lower addition of
mass to the lake and resulting in an enthalpy value that is double than expected
for saturated steam at hydrostatic pressures likely to be found at the bottom of
the lake. By including uncertainties in our observations, this explanation is
still a possibility but no longer a requirement to satisfy the physical
constraints. Numerical modelling of hydrothermal processes may also shed further
light on the heating processes and whether or not the lake and the vent form a
single or two separate convective systems.
